Finding and controling biases in dark matter halo concentration estimates

### 2016-07-08 paper/ms.tex
nueva version completa

\begin{abstract}
We use bootstrapping to estimate the bias of concentration
estimates on N-body dark matter halos as a function of particle
number. 
We find that algorithms based on the maximum radial velocity
and radial particle binning have increaing biases with decreasing 
particle numbers.
These methods overestimate the concentration by 
$15\%-20\%$ for halos sampled with $200$ particles, and $7\%$-$8\%$
for halos sampled with $500$ particles. 
To control this bias level we propose a new algorithm based on the
integrated mass profile. 
The method uses the full particle information without any binning,
making it reliable in cases when low numerical resolution becomes a
limitation for other methods.
This method reduces the bias to $2\%-3\%$ for halos sampled with
$200$-$500$ particles.
To keep the biases of the velocity and density methods to the same
low levels as in the new method, only halos sampled
with at least $\sim 4000$ particles should be used.
We also show that the mass-concentration relationship could be
shallower than expected from numerical experiments once the biases 
of the different concentration measurements are taken into account.
From these results we believe that bootstrapping and the
concentration estimation based on the integrated mass are
valuable tools to probe the internal structure of dark matter halos
in numerical simulations.
\end{abstract}

@@

@@

We notice first that the results from the density method have a
systematic $15\%$offset from the velocity methods.  
This offset was already presented by \citep{Prada2012} for low
concentrations ($c<6$) and high ($M_h>10^{12}\hMsun$) halo masses.  
Recently \citep{Klypin2016} summarized results for the
mass-concentration relationship coming from different methods and
datasets to show that similar systematic offsets are present.
\citep{2014MNRAS.441.3359D} studied the mass-concentration
relationship using the maximum velocity and density methods and did
not report any significant difference. However, they implemented a
modified version of the velocity algorithm that uses the data binning
obtained for the density algorithm.

The new method closely follows the results of
the velocity method at higher masses, $M_h>10^{12}\hMsun$ or
equivalently for $>5\times10^3$ particles.  
For masses below that threshold the new method saturates at a
concentration value of $\sim 8$ while the velocity method continues to
give higher concentration values.

We hypothesize that the increase in the results for the velocity
method below $5\times 10^{4}$ particles come from the systematic bias
described in the previous subsection.  To test the general consistency
of this hypothesis we correct the concentration values in the
velocity and integrated mass methods by a factor of $1/(1+f_{\rm
  off})$, using the definition in Equation (\ref{eq:f_off}) and the 
parameters obtained from the data presented in Figure
(\ref{fig:downsampling}).  
The correction brings into perfect agreement the results between the
velocity and the integrated mass method.  
This suggests that the concentration at lower masses might very well
reach a plateau in the halo mass range $10^{10}< M_{h}/\hMsun <
10^{12}$ .

@@

In this paper we quantified biases on concentration numerical estimates by
bootstrapping dark matter halos resolved with a  large number of
particles ($\sim 10^5$).
We found that methods commonly used in the literature present a large
bias by overestimating concentrations by factors of $13\%$-$20\%$ for
halos with $200$ particles, or $7\%$-$10\%$ for halos with $500$
particles.  
This procedure is a new way to quantify the bias in concentration
estimates with the advantage that it works without having to run new
simulations.

>>>>

\begin{abstract}
We use bootstrapping to estimate the bias of concentration
estimates on N-body dark matter halos as a function of particle
number. 
We find that algorithms based on the maximum radial velocity
and radial particle binning have increaing biases with decreasing 
particle numbers.
These methods overestimate the concentration by 
$15\%-20\%$ for halos sampled with $200$ particles, and $7\%$-$10\%$
for halos sampled with $500$ particles. 
To control this bias level we propose a new algorithm based on the
integrated mass profile. 
The method uses the full particle information without any binning,
making it reliable in cases when low numerical resolution becomes a
limitation for other methods.
This method reduces the bias to $2\%-3\%$ for halos sampled with
$200$-$500$ particles.
To keep the biases of the velocity and density methods to the same
low levels as in the new method, only halos sampled
with at least $\sim 4000$ particles should be used.
We also show that the mass-concentration relationship could be
shallower than expected from numerical experiments once the biases 
of the different concentration measurements are taken into account.
From these results we believe that bootstrapping and the
concentration estimation based on the integrated mass are
valuable tools to probe the internal structure of dark matter halos
in numerical simulations.
\end{abstract}

@@

We now turn to the question of quantifying the concentration results
of the new algorithm and possible applications of the biases found.

@@

We notice first that the results from the new algorithm follow very
closely the velocity algorithm at high masses ($M_h>10^{12}\hMsun$ or
equivalently for $>8\times10^3$ particles). 
For lower masses there is a difference between the median of the two
methods, but they are still consistent within the statistical
uncertainties.

We also notice that the results from the density method have a
systematic $15\%$offset from the velocity methods.  
This offset was already presented by \cite{Prada2012} for low
concentrations ($c<6$) and high ($M_h>10^{12}\hMsun$) halo masses.  
Recently \citep{Klypin2016} summarized results for the
mass-concentration relationship coming from different methods and
datasets to show that similar systematic offsets are present.
\citep{2014MNRAS.441.3359D} studied the mass-concentration
relationship using the maximum velocity and density methods and did
not report any significant difference. 
However, they implemented a modified version of the velocity algorithm
that uses the data binning obtained for the density algorithm.

We hypothesize that the increase in the results for the velocity
and density methods below $8\times 10^{3}$ particles come from the
systematic bias described in the previous section.  
To test the general consistency of this hypothesis we correct the
concentration values in the velocity and integrated mass methods by a
factor of $1/(1+f_{\rm  off})$, using the definition in Equation
(\ref{eq:f_off}) and the  
parameters obtained from the data presented in Figure
(\ref{fig:downsampling}).  
The correction brings into perfect agreement the results between the
velocity/density methods and the new algorithm.

@@

In this paper we quantified biases on concentration numerical estimates by
bootstrapping dark matter halos resolved with a  large number of
particles ($\sim 10^5$).
We found that methods commonly used in the literature present a large
bias by overestimating concentrations by factors of $15\%$-$20\%$ for
halos with $200$ particles, or $7\%$-$10\%$ for halos with $500$
particles.  
This procedure is a new way to quantify the bias in concentration
estimates with the advantage that it works without having to run new
simulations.
